---
title: "Final Project: Online Sales Analysis "
authors: "Khushi Satra, 203383100
          Satinder Kaur, 169109308
          Sara Haifa, 169087012
          Spencer Mozeg , 169099531
          Noya Barak, 
---


#Load packages
```{r}
library(tidyverse)
library(tidymodels)
library(ggplot2)
```
#Load datasets
```{r}
Movie_industry<- read.csv("https://raw.githubusercontent.com/khushisatra/Final-Project/refs/heads/main/movies.csv")
Movie_industry

movies_industry_data$rating[movies_industry_data$rating == "" ] <- NA

```
{r}
movie_industry <- na.omit(Movie_industry)
movie_industry

#Step 1: Tidying 
```{r}
clean_movie_industry <- movie_industry|>
  mutate(
  released_date = str_extract(released, "^[A-Za-z]+ \\d{1,2}, \\d{4}"),
  hours = runtime%/%60,             
    minutes = runtime %% 60, 
    movie_duration = sprintf("%d hours, %d minutes", hours, minutes),
  rating = ifelse(rating == "X", "NC-17",
                         ifelse(rating == "Approved", "G", rating)),
   Profit = gross - budget
  ) |>
  select(-released, -year, -hours, -minutes, -runtime)|>
  filter(rating != "Not Rated" & rating != "Unrated")|>
   mutate(genre = as.factor(genre),
  rating = as.factor(rating))|>
  rename(
    IMDb_score = score,
    user_votes = votes,
    gross_revenue = gross
  )
  
clean_movie_industry

```
#Exploratory Plot 1 - Visualize Average Profit and Revenue By Genre

profit_revenue <- clean_movie_industry|>
  group_by(genre)|>
  summarise(avg_Profit = mean(Profit, na.rm = TRUE),
            avg_revenue = mean(gross_revenue, na.rm = TRUE)) %>%
  pivot_longer(cols = c(avg_Profit, avg_revenue), 
               names_to = "metric", 
               values_to = "value")

ggplot(profit_revenue, aes(x = reorder(genre, value), y = value, fill = metric)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_grid(rows = vars(metric))+
  coord_flip() +
  labs(title = "Comparison of Profit and Revenue by Genre",
       x = "Genre", y = "Value", fill = "Metric") +
  theme_minimal()

#Exploratory Plot 2 - Compare IMDb scores and profits across genres

genre_summary <- clean_movie_industry|>
  group_by(genre) |>
  summarise(avg_Profit = mean(Profit))|>
  arrange(desc(avg_Profit))

genre_comparison <- clean_movie_industry|>
  left_join(genre_summary, by = "genre" )

ggplot(genre_comparison, aes(x = reorder(genre, avg_Profit), y = IMDb_score, fill = avg_Profit)) +
  geom_boxplot(outlier.color = "red", outlier.shape = 16, outlier.size = 2) +
  coord_flip() +
  scale_fill_gradient(low = "lightblue", high = "darkblue", name = "Avg Profit") +
  labs(title = "Distribution of IMDb Scores by Genre", 
       x = "Genre (Ordered by Avg Profit)", 
       y = "IMDb Score") +
  theme_minimal()

#Spliting 
```
clean_movie_industry_split <- initial_validation_split(clean_movie_industry, prop = c(0.6,0.2)) 
  
  clean_movie_ind_val <- validation(clean_movie_industry_split)
  clean_movie_ind_train <- training(clean_movie_industry_split)
  clean_movie_ind_test <- testing(clean_movie_industry_split)
  
clean_movie_industry_split
```

#Recipes
```
recipe1 <- recipe(Profit ~ budget + genre, data =  clean_movie_ind_train) |> 
  step_dummy(all_nominal_predictors()) 

recipe1 #Higher budgets genrally lead to higher profitability; look at how this spreads across genres. 


recipe2 <- recipe(Profit ~ IMDb_score + rating, data = clean_movie_ind_train) |>
  step_dummy(all_nominal_predictors()) 

recipe2 #Higher profits are associated with certain ratings, accoutning quality with IDMB_score


clean_movie_ind_recipes <- list(
  genre_profit_analysis= recipe1,
  profit_score = recipe2
  )

clean_movie_ind_recipes
```
