---
title: "Final Project: Online Sales Analysis "
authors: "Khushi Satra, 203383100
          Satinder Kaur, 169109308
          Sara Haifa, 169087012
          Spencer Mozeg , 169099531
          Noya Barak, 
---


#Load packages
```{r}
library(tidyverse)
library(ggrepel)
theme_set(theme_bw())
library(arrow)
library(tidymodels)
tidymodels_prefer()
```
#Load datasets
```{r}
movie_industry<- read.csv("https://raw.githubusercontent.com/khushisatra/Final-Project/refs/heads/main/movies.csv")

movie_industry$rating[movie_industry$rating == "" ] <- NA
movie_industry <- na.omit(movie_industry)
```


#Step 1: Tidying 
```{r}

clean_movie_industry <- movie_industry|>
  mutate(
  released_date = str_extract(released, "^[A-Za-z]+ \\d{1,2}, \\d{4}"),
  hours = runtime%/%60,             
    minutes = runtime %% 60, 
    movie_duration = sprintf("%d hours, %d minutes", hours, minutes),
  rating = ifelse(rating == "X", "NC-17",
                         ifelse(rating == "Approved", "G", rating)),
   Profit = gross - budget
  ) |>
  select(-released, -year, -hours, -minutes, -runtime)|>
  filter(rating != "Not Rated" & rating != "Unrated")|>
   mutate(genre = as.factor(genre),
  rating = as.factor(rating))|>
  mutate(
    budget = round(budget/1000000, 2),
    Profit = round(Profit/1000000, 2),
    gross = round(gross/1000000,2)
  )|>
  rename(
    IMDb_score = score,
    user_votes = votes,
    gross_revenue_mil = gross,
    profit_mil = Profit,
    budget_mil = budget
  )
  
clean_movie_industry

```

#Exploratory Plot 1 - Visualize Average Profit and Revenue By Genre
```{r}
profit_revenue <- clean_movie_industry|>
  group_by(genre)|>
  summarise(avg_profit = mean(profit_mil, na.rm = TRUE),
            avg_revenue = mean(gross_revenue_mil, na.rm = TRUE))|>
  pivot_longer(cols = c(avg_profit, avg_revenue), 
               names_to = "metric", 
               values_to = "value")

ggplot(profit_revenue, aes(x = reorder(genre, value), y = value, fill = metric)) +
  geom_bar(stat = "identity", position = "dodge") +
  coord_flip() +
  labs(title = "Comparison of Profit and Revenue by Genre",
       x = "Genre", y = "Value in MIllions ($)", fill = "Metric",
       caption = "Source: movies_industry dataset") +
  theme_minimal()
```

#Exploratory Plot 2 - Compare IMDb scores and profits across genres
```{r}
genre_summary <- clean_movie_industry|>
  group_by(genre) |>
  summarise(avg_profit = mean(profit_mil))|>
  arrange(desc(avg_profit))

genre_comparison <- clean_movie_industry|>
  left_join(genre_summary, by = "genre" )

ggplot(genre_comparison, aes(x = reorder(genre, avg_profit), y = IMDb_score, fill = avg_profit)) +
  geom_boxplot(outlier.color = "red", outlier.shape = 16, outlier.size = 2) +
  coord_flip() +
  scale_fill_gradient(low = "lightblue", high = "darkblue", name = "Avg Profit in Million ($)") +
  labs(title = "Distribution of IMDb Scores by Genre", 
       subtitle = "Coloured by Average Profit in Millions ($)",
       x = "Genre (Ordered by Avg Profit)", 
       y = "IMDb Score") +
  theme_minimal()
```

#Spliting 
```{r}
clean_movie_industry_split <- initial_validation_split(clean_movie_industry, prop = c(0.6,0.2)) 
  
  clean_movie_ind_val <- validation(clean_movie_industry_split)
  clean_movie_ind_train <- training(clean_movie_industry_split)
  clean_movie_ind_test <- testing(clean_movie_industry_split)
  
clean_movie_industry_split
```

#Recipes
```{r}
basic_recipe1 <- recipe(profit_mil ~ gross_revenue_mil + rating +genre + country , data =  clean_movie_ind_train)|> 
    step_dummy(all_nominal_predictors()) 

interaction_recipe2 <- recipe(profit_mil ~ genre + country + IMDb_score + user_votes + budget_mil, data = clean_movie_ind_train) |>
  step_interact(~ IMDb_score:genre) |>
  step_log(IMDb_score, base = 10) |>
  step_normalize(all_numeric_predictors(), -all_outcomes())|>
  step_dummy(all_nominal_predictors())


clean_movie_ind_recipes <- list(
  recipe1 = basic_recipe1,
  recipe2 = interaction_recipe2
  )

clean_movie_ind_recipes
```
#Linear Model 1 
```{r}
ggplot(clean_movie_ind_train, aes(x = gross_revenue_mil, y = profit_mil, colour = rating)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "blue")+
  labs(
    title = "Interaction between Gross Revenue and Profit",
    x = "Gross Revenue in Million ($)",
    y = "Profit in Million ($)"
  )
```

#Linear Model 2
```{r}
ggplot(clean_movie_ind_train, aes(x = IMDb_score, y = profit_mil, color = genre)) +
    geom_point(alpha = 0.6) +
    geom_smooth(method = "lm", se= FALSE, color = "blue") +
    labs(title = "Profit vs IMDb Score by Genre", x = "IMDb Score", y = "Profit") +
    theme_minimal()
```

#Workflow Creation
```{r}
movies_workflow_set <- workflow_set(clean_movie_ind_recipes,models = list(lm= linear_reg()), cross = FALSE)

movies_workflow_set
```
#RMSE Calculation
```{r}
set.seed(0474100)

movies_fit <- movies_workflow_set |>
  workflow_map(fn = "tune_grid",
               seed = 100,
               resamples = validation_set(clean_movie_industry_split))

movies_fit

# Plot for results
movies_fit |>
    collect_metrics() |>
    ggplot() +
    aes(x = wflow_id, y = mean) +
    geom_col(fill = "lightgrey", colour = "black") +
    facet_wrap(~ .metric, scales = "free")
```

#Reduced Features
```{r}
reduced_basic_recipe1 <- recipe(profit_mil ~ gross_revenue_mil + genre + country , data =  clean_movie_ind_train)|> #removed rating
    step_dummy(all_nominal_predictors()) 

reduced_interaction_recipe2 <- recipe(profit_mil ~ genre + IMDb_score + budget_mil, data = clean_movie_ind_train) |> # removed user_votes, country
  step_interact(~ IMDb_score:genre) |>
  step_log(IMDb_score, base = 10) |>
  step_normalize(all_numeric_predictors(), -all_outcomes())|>
  step_dummy(all_nominal_predictors())


reduced_movie_recipes <- list(
  reduced_recipe1 = reduced_basic_recipe1,
  reduced_recipe2 = reduced_interaction_recipe2
  )

reduced_movie_recipes
```

#Workflow for reduced features
```{r}
reduced_movies_workflow_set <- workflow_set(reduced_movie_recipes,models = list(lm= linear_reg()), cross = FALSE)

reduced_movies_workflow_set
```

#RMSE for reduced feature
```{r}
set.seed(0474100)

reduced_movies_fit <- reduced_movies_workflow_set |>
  workflow_map(fn = "tune_grid",
               seed = 100,
               resamples = validation_set(clean_movie_industry_split))

reduced_movies_fit

#Plot for Results
reduced_movies_fit |>
    collect_metrics() |>
    ggplot() +
    aes(x = wflow_id, y = mean) +
    geom_col(fill = "lightgrey", colour = "black") +
    facet_wrap(~ .metric, scales = "free")

```

#Advanced Modelling
```{r}
train_test <- initial_split(clean_movie_industry, prop = 0.8)

rf <- rand_forest(trees = 1000, mtry = tune(), min_n = 3) |>
    set_engine("randomForest") |>
    set_mode("regression")
tune_grid <- tibble(mtry = 1:3)

lm_versus_rf <- workflow_set(
  preproc = list(
    recipe1 = basic_recipe1,
    recipe2 = interaction_recipe2
  ),
  models = list(lm= linear_reg(), rf = rf)
  ) |>
    workflow_map(
        fn = "tune_grid",
        grid = tune_grid,
        seed = 100,
        control = control_grid(save_pred = TRUE, save_workflow = TRUE),
        resamples = vfold_cv(training(clean_movie_industry_split), v = 5)
    )

lm_versus_rf |>
    autoplot(select_best = TRUE, metric = "rmse") +
    geom_label_repel(aes(label = wflow_id))

```

#Best Model Decision
```{r}

best_lm_workflow <- lm_versus_rf |>
    extract_workflow_set_result("recipe1_lm") |>
    select_best(metric = "rmse") 

best_lm <- lm_versus_rf |>
  extract_workflow("recipe1_lm") |>
  finalize_workflow(best_lm_workflow)

best_lm

best_rf_workflow <- lm_versus_rf |>
  extract_workflow_set_result("recipe1_rf") |>
  select_best(metric = "rmse")

best_rf <- lm_versus_rf |>
  extract_workflow("recipe1_rf") |>
  finalize_workflow(best_rf_workflow)

best_rf
```

#Test fit
```{r}
test_lm <- last_fit(best_lm, split = clean_movie_industry_split)

test_rf <- last_fit(best_rf, split = clean_movie_industry_split)

cat("\nBest RMSE and R^2:\n")
bind_rows(
    lm = collect_metrics(test_lm),
    rf = collect_metrics(test_rf),
    .id = "model"
)

cat("\nPredictors used:\n")
extract_recipe(test_lm) |> formula()
extract_recipe(test_rf) |> formula()

# Plot the predictions versus observed
bind_rows(
    lm = collect_predictions(test_lm),
    rf = collect_predictions(test_rf),
    .id = "model"
) |>
    ggplot() +
    aes(x = Profit, y = .pred, colour = model) +
    geom_point(shape = 1)

bind_rows(
    lm = test_lm,
    rf = test_rf,
    .id = "model"
) |>
    # This is almost certainly a better way to do this, but I do not know it.
    unnest_wider(.metrics) |>
    unnest_longer(c(.metric, .estimator, .estimate, .config)) |>
    ggplot() +
    aes(x = model, y = .estimate) +
    geom_col(fill = "lightgrey", color = 1) +
    facet_wrap(~ .metric, scales = "free_y")
```





