---
title: "Final Project: Online Sales Analysis "
authors: "Khushi Satra, 203383100
          Satinder Kaur, 169109308
          Sara Haifa, 169087012
          Spencer Mozeg , 169099531
          Noya Barak, 169097527"
---
Introduction 
This project looks into the online sale trends in the movie industry, targeting key factors like earnings, generating revenue, and consumers being received. Using strong R measures such as tidyverse and ggplot2, the project examines data patterns to gain knowledge into what drives business success. In order to look into the correlations among variables like movie genres, fabrication  of budgets, income, and audience ratings, the project involves organizing and arranging the data, producing data visualizations, and building statistical models. Our initial understanding leads us to the theory that genres or categories with a moderate production costs and higher audience assessments are more likely to be profitable.  For manufacturers, businesses, and the filmmakers looking for ways to optimize their funding, investments and focus on the most profitable segments of the film industry, this expected conclusion is meant to offer insightful information.

Goals/Research Question 
The goal of this analysis is to respond to the following research question: which variables affect the earnings of films, and what is the correlation between consumers ratings and genres and their critical and economic success? This project aims to address this by examining the Internet Movie Database score payments, contrasting typical earnings and sales by genre, and forecasting profit using attributes like income and assessment. The necessary packages (tidyverse, tidymodels, ggplot2) are installed, the dataset is cleaned, and then modifications are used to generate new variables like earnings, generate movie durations, and establish a consistent format. For clarity, variables are renamed and insignificant ratings have been removed. Examining the Internet Movie Database score variations across genres and displaying typical profit and revenue by genre are examples of exploratory research that highlights consumers and economic patterns. Using trend lines and scatter plots, still early demonstrating examines the connection between revenue and profit. In the future, recipes will be enhanced, new models will be created, and their effectiveness will be evaluated using figures like R2 or RMSE. Further insights into the revenues and public reaction patterns of the industry may be obtained through extra research, such as graphs of correlation or time based trends. This workflow provides an all encompassing method for comprehending and forecasting significant variables in the film industry.

---

Load packages
```{r}
library(tidyverse)
library(ggrepel)
theme_set(theme_bw())
library(arrow)
library(tidymodels)
tidymodels_prefer()
```
Load datasets
```{r}
movie_industry<- read.csv("https://raw.githubusercontent.com/khushisatra/Final-Project/refs/heads/main/movies.csv")
```

Tidying 
To clean this data set, we want to make sure that we removed all non-applicable values to ensure consistency and more accurate calculations - this includes values that are unknown, missing, and empty. We also standardized the movie ratings to give a clearer definition, and amplified the movie's duration. Furthermore, we were curious in the profit of these movies, and any relation with other features complementing profit; therefore we calculated said profit for each movie using its gross and budget data. The net profit for a movie indicates whether it made a profit or incurred a loss, with a positive value indicating more profit than production costs, which will become our key target variable throughout this report.

```{r}
movie_industry$rating[movie_industry$rating == "" ] <- NA
movie_industry <- na.omit(movie_industry)

clean_movie_industry <- movie_industry|>
  mutate(
  released_date = str_extract(released, "^[A-Za-z]+ \\d{1,2}, \\d{4}"),
  hours = runtime%/%60,             
    minutes = runtime %% 60, 
    movie_duration = sprintf("%d hours, %d minutes", hours, minutes),
  rating = ifelse(rating == "X", "NC-17",
                         ifelse(rating == "Approved", "G", rating)),
   Profit = gross - budget
  ) |>
  select(-released, -year, -hours, -minutes, -runtime)|>
  filter(rating != "Not Rated" & rating != "Unrated")|>
   mutate(genre = as.factor(genre),
  rating = as.factor(rating))|>
mutate(
budget = round(budget/1000000, 2),
Profit = round(Profit/1000000, 2),
gross = round(gross/1000000,2)
)|>
  rename(
    IMDb_score = score,
    user_votes = votes,
    gross_revenue = gross,
    gross_revenue_mil = gross,
     profit_mil = profit,
      budget_mil = budget
  )
  
clean_movie_industry

```

Exploratory Plot #1 - Visualize Average Profit and Revenue By Genre
```{r}
profit_revenue <- clean_movie_industry|>
  group_by(genre)|>
  summarise(avg_Profit = mean(Profit, na.rm = TRUE),
            avg_revenue = mean(gross_revenue, na.rm = TRUE)) %>%
  pivot_longer(cols = c(avg_Profit, avg_revenue), 
               names_to = "metric", 
               values_to = "value")

ggplot(profit_revenue, aes(x = reorder(genre, value), y = value/1e6, fill = metric)) +
  geom_bar(stat = "identity", position = "dodge") +
  #facet_grid(rows = vars(metric))+
  coord_flip() +
  labs(title = "Comparison of Profit and Revenue by Genre",
       x = "Genre", y = "Value in Millions ($)", fill = "Metric",
       caption = "Source: movies_industry dataset") +
  theme_minimal()
```

Exploratory Plot #2 - Compare IMDb scores and profits across genres
```{r}
genre_summary <- clean_movie_industry|>
  group_by(genre) |>
  summarise(avg_Profit = mean(Profit))|>
  arrange(desc(avg_Profit))

genre_comparison <- clean_movie_industry|>
  left_join(genre_summary, by = "genre" )

ggplot(genre_comparison, aes(x = reorder(genre, avg_Profit), y = IMDb_score, fill = avg_Profit/1e6)) +
  geom_boxplot(outlier.color = "red", outlier.shape = 16, outlier.size = 2) +
  coord_flip() +
  scale_fill_gradient(low = "lightblue", high = "darkblue", name = "Avg Profit in Million ($)") +
  labs(title = "Distribution of IMDb Scores by Genre", 
       subtitle = "Coloured by Average Profit in Millions ($)",
       x = "Genre (Ordered by Avg Profit)", 
       y = "IMDb Score") +
  theme_minimal()
```

Spliting 
Before modelling, we want to split the data to properly prepare for evaluation. By splitting the data into these three sets before applying the recipes, we ensure that the models are not overfitting to the training data and can generalize well to new data. 
```{r}
clean_movie_industry_split <- initial_validation_split(clean_movie_industry, prop = c(0.6,0.2)) 
  
  clean_movie_ind_val <- validation(clean_movie_industry_split)
  clean_movie_ind_train <- training(clean_movie_industry_split)
  clean_movie_ind_test <- testing(clean_movie_industry_split)
  
clean_movie_industry_split
```

Recipes
The recipes are utilized to preprocess data, creating dummy variables for categorical predictors and incorporating interaction terms, preparing it for the models furthermore and capturing its relationships.
```{r}
basic_recipe1 <- recipe(Profit ~ gross_revenue + rating, data =  clean_movie_ind_train)|> 
    step_dummy(all_nominal_predictors()) 

interaction_recipe2 <- recipe(Profit ~ genre + country + IMDb_score + user_votes + budget, data = clean_movie_ind_train) |>
  step_interact(~ IMDb_score:genre) 


clean_movie_ind_recipes <- list(
  recipe1 = basic_recipe1,
  recipe2 = interaction_recipe2
  )

clean_movie_ind_recipes
```
Linear Model 1 - Relationship Between Gross Revenue and Profit by Movie Rating: 
This linear model examines the relationship between gross revenue and profit for movies, revealing how revenue predicts profitability and how movie ratings influence this relationship. 

The scatterplot shows individual movies as points, with gross revenue on the x-axis, profit on the y-axis, and ratings distinguished by color. The plot reveals a strong positive linear relationship between gross revenue and profit, indicating that higher revenue consistently leads to increased profitability across movies. Ratings appear to have minimal impact on this trend, with most movies aligning closely to the trendline regardless of their category. Outliers, such as blockbuster films (specifically PG-13 rating) with exceptionally high revenue and profit, and niche categories like TV-MA movies with lower performance, provide areas for further investigation.This analysis is crucial for understanding financial dynamics, identifying cost inefficiencies, and optimizing business decisions in movie production.

In fact, according to Axios2,  an analysis of box office performance by movie ratings from 1995 to 2022 shows that PG-13 films generally outperform movies with both higher and lower ratings. G-rated films tend to perform the worst, followed by PG and then R-rated movies. Statista3 reports that in 2022, PG-13 movies made up approximately two-thirds of U.S. box office revenue.

```{r}
ggplot(clean_movie_ind_train, aes(x = gross_revenue/1e6 , y = (Profit/1e6), colour = rating)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "blue")+
  labs(
    title = "Interaction between Gross Revenue and Profit",
    x = "Gross Revenue in Million ($)",
    y = "Profit in Million ($)"
  )
```

#Linear Model 2
```{r}
ggplot(clean_movie_ind_train, aes(x = IMDb_score * as.numeric(genre), y = Profit/1e6)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Interaction between IMDb Score and Genre vs Profit",
    x = "IMDb Score * Genre (Interaction)",
    y = "Profit in Million ($)"
  )

```
```{r}
movies_workflow_set <- workflow_set(clean_movie_ind_recipes,models = list(lm= linear_reg()), cross = FALSE)

movies_workflow_set
```

```{r}
set.seed(0474100)

movies_fit <- movies_workflow_set |>
  workflow_map(fn = "tune_grid",
               seed = 100,
               resamples = validation_set(clean_movie_industry_split))

movies_fit

Plot for Results
movies_fit |>
    collect_metrics() |>
    ggplot() +
    aes(x = wflow_id, y = mean) +
    geom_col(fill = "lightgrey", colour = "black") +
    facet_wrap(~ .metric, scales = "free")
```
Advanced Modelling
```{r}
train_test <- initial_split(clean_movie_industry, prop = 0.8)

rf <- rand_forest(trees = 1000, mtry = tune(), min_n = 3) |>
    set_engine("randomForest") |>
    set_mode("regression")
tune_grid <- tibble(mtry = 1:3)

lm_versus_rf <- workflow_set(
  preproc = list(
    recipe1 = basic_recipe1,
    recipe2 = interaction_recipe2
  ),
  models = list(lm= linear_reg(), rf = rf)
  ) |>
    workflow_map(
        fn = "tune_grid",
        grid = tune_grid,
        seed = 100,
        control = control_grid(save_pred = TRUE, save_workflow = TRUE),
        resamples = vfold_cv(training(clean_movie_industry_split), v = 5)
    )

lm_versus_rf |>
    autoplot(select_best = TRUE, metric = "rmse") +
    geom_label_repel(aes(label = wflow_id))

```
```{r}

best_lm_workflow <- lm_versus_rf |>
    extract_workflow_set_result("recipe1_lm") |>
    select_best(metric = "rmse") 

best_lm <- lm_versus_rf |>
  extract_workflow("recipe1_lm") |>
  finalize_workflow(best_lm_workflow)

best_lm

best_rf_workflow <- lm_versus_rf |>
  extract_workflow_set_result("recipe1_rf") |>
  select_best(metric = "rmse")

best_rf <- lm_versus_rf |>
  extract_workflow("recipe1_rf") |>
  finalize_workflow(best_rf_workflow)

best_rf
```
#test fit
```{r}
test_lm <- last_fit(best_lm, split = clean_movie_industry_split)

test_rf <- last_fit(best_rf, split = clean_movie_industry_split)

cat("\nBest RMSE and R^2:\n")
bind_rows(
    lm = collect_metrics(test_lm),
    rf = collect_metrics(test_rf),
    .id = "model"
)

cat("\nPredictors used:\n")
extract_recipe(test_lm) |> formula()
extract_recipe(test_rf) |> formula()

# Plot the predictions versus observed
bind_rows(
    lm = collect_predictions(test_lm),
    rf = collect_predictions(test_rf),
    .id = "model"
) |>
    ggplot() +
    aes(x = Profit, y = .pred, colour = model) +
    geom_point(shape = 1)

bind_rows(
    lm = test_lm,
    rf = test_rf,
    .id = "model"
) |>
    # This is almost certainly a better way to do this, but I do not know it.
    unnest_wider(.metrics) |>
    unnest_longer(c(.metric, .estimator, .estimate, .config)) |>
    ggplot() +
    aes(x = model, y = .estimate) +
    geom_col(fill = "lightgrey", color = 1) +
    facet_wrap(~ .metric, scales = "free_y")
```
---
Conclusion
This analysis wanted to determine the main factors influencing movie earnings, with an emphasis on the relationship between genres , categories, consumers ratings, and economic success. According to the results, genres with equal production costs and larger Internet Movie Database scores tend to be more successful. People in the industry may find these results useful in improving their marketing strategies and choices. Using the random forest and linear regression models, we were able estimate profitability based on key factors such as budget, revenue from advertising, and Internet Movie Database ratings. In this case, nevertheless, the linear model surpassed the random forest model, pointing to that sometimes simpler methods could yield more precise predictions. In addition, the model that anticipated profit the best was the one based on Recipe 1. All things taken into account, this evaluation provides helpful information into the ways that film characteristics affect revenue, setting up the foundation for future initiatives and further research into the shifting circumstances associated to the motion picture sector.

Limitations
There are several restrictions on the code. The data cleaning procedure may introduce unfairness because it assumes that missing assessments should be set to `NA` or specific groups. Using `na.omit()` could result in the loss of significant data without dealing with the causes of missing values. It is unclear why certain feature development choices such as converting ratings and genres to variables and creating the `movie_duration` variable, were made. The outliers are not taken into account in the exploratory plots, which could skew the findings. For high values, the linear model's reliance on linear relationships might not hold true. Additional hyperparameter, which adjustment and verification must be done for the random forest model. In addition, the validation method lacks strong crossover validation and depends on an easily understood split. Plot design and scaling problems could also be fixed. The reliability and comprehension of the analysis could be improved by addressing these limitations.
---

References 
1. Movie Industry (Daniel Grijalva, updated 3 years ago): https://www.kaggle.com/datasets/danielgrijalvas/movies 

2. R-rated movies hit lowest box office share in more than 25 years (Sara Fischer, 2022) : https://www.axios.com/2022/08/16/r-rate-movies-box-office 

3. U.S. & Canada: PG-13 films' box office market share 2023 | Statista: https://www.statista.com/statistics/792505/domestic-box-office-pg-13-movies-market-share/



